---
title: "WEEK 3  NYPD Shooting Incident Data Report"
author: "HDaher"
date: "2024-04-06"
output:
  html_document: default
  pdf_document: default
---

------------------------------------------------------------------------

# Description

-   Data set of every Shooting Incident in NYC since 2006 to end of previous calendar year.
-   Each record includes information about the event, the location and time of occurrence.
-   information related to suspect and victim demographics is also included.
-   This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website.

```{r}
url_NYPD <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

NYPD <- read.csv(url_NYPD)

```

# Libraries

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(broom)
```

# Data Exploration

```{r}
dim(NYPD)
tail(NYPD)
#numerical statistics summary
summary(NYPD[, sapply(NYPD, is.numeric)])
#categorical statistics summary
summary(NYPD[, sapply(NYPD, is.factor)])
# Looking at column values
glimpse(NYPD)
names(NYPD)

```

# Data Preparation

```{r}
# Convert categorical variables to factors
NYPD <- NYPD %>%
  mutate_at(vars(BORO, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE), factor)
# Remove unnecessary columns 
NYPD <- NYPD %>% select(-Lon_Lat)

summary(NYPD)

sapply(NYPD, function(x) sum(is.na(x)))

names(NYPD)

```

# Exploring Incident Time Series Data

## Total Yearly Incidents

```{r}
# Change OCCUR_DATE to Date format
NYPD$OCCUR_DATE <- as.Date(NYPD$OCCUR_DATE, format = "%m/%d/%Y")

# Get year
NYPD$YEAR <- format(NYPD$OCCUR_DATE, "%Y")

# Group by year
yearly_incidents <- NYPD %>%
  group_by(YEAR) %>%
  summarise(Incidents = n())

# Plot total yearly incidents 
g <- ggplot(yearly_incidents, aes(x = YEAR, y = Incidents)) +
  geom_col(fill = "#3BB9FF", colour = "black") +  
  labs(title = "Total Yearly Incidents in NYC",
       x = "Year",
       y = "Number of Incidents") +
  theme(legend.position = "none")

# Print the plot
print(g)

```

## Total Daily Incidents

```{r}

# Get day
NYPD$DAY_OF_WEEK <- weekdays(NYPD$OCCUR_DATE)

# Group totals by day of week
daily_incidents <- NYPD %>%
  group_by(DAY_OF_WEEK) %>%
  summarise(Incidents = n())


order_days <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

# Plot total daily incidents 
g <- ggplot(daily_incidents, aes(x = factor(DAY_OF_WEEK, levels = order_days), y = Incidents)) +
  geom_col(fill = "#3BB9FF", color = "black") + 
  labs(title = "Total Incidents by Day of the Week in NYC",
       x = "Day of the Week",
       y = "Number of Incidents") +
  theme(legend.position = "none")

# Print the plot
print(g)


```

```{r}
# Extract hour of the day from OCCUR_TIME
NYPD$HOUR_OF_DAY <- as.numeric(format(strptime(NYPD$OCCUR_TIME, format = "%H:%M:%S"), "%H"))

# Group data by hour of the day and count the number of incidents
hourly_incidents <- NYPD %>%
  group_by(HOUR_OF_DAY) %>%
  summarise(Incidents = n())


# Plot total incidents vs hour of day
g <- ggplot(hourly_incidents, aes(x = HOUR_OF_DAY, y = Incidents)) +
  geom_point(color = "#3BB9FF") +
  geom_smooth(method = "loess", se = FALSE, color = "#5EFB6E") +
  labs(title = "Total Incidents by Hour of the Day in NYC",
       x = "Hour of the Day",
       y = "Number of Incidents") 


print(g)


```

# Exploring Location Data

## Total Incidents by Borough

```{r}
# Group data by BORO
boro_incidents <- NYPD %>%
  group_by(BORO) %>%
  summarise(Incidents = n())

# Plot total incidents by BORO
g <- ggplot(boro_incidents, aes(x = reorder(BORO, -Incidents), y = Incidents)) +
  geom_bar(stat = "identity", fill = "#3BB9FF", color = "black") +  
  labs(title = "Total Incidents by Borough in NYC",
       x = "Borough",
       y = "Number of Incidents") +
  coord_flip()  


print(g)

```

## Total Incidents by Location Classification

```{r}
# Create a copy of the NYPD dataset
NYPD_copy <- NYPD

# Filter out blanks and NAs, then summarize incidents by location classification
classfctn_incidents <- NYPD_copy %>%
  filter(LOCATION_DESC != "", !is.na(LOCATION_DESC)) %>%
  group_by(LOCATION_DESC) %>%
  summarise(Incidents = n()) %>%
  arrange(desc(Incidents)) %>%
  top_n(10)  # Select top 10 

# Plot total incidents for the top 10 location classifications
g <- ggplot(classfctn_incidents, aes(x = reorder(LOCATION_DESC, -Incidents), y = Incidents)) +
  geom_bar(stat = "identity", fill = "#3BB9FF", color = "black") +
  labs(title = "Top 10 Locations by Total Incidents in NYC ",
       x = "Location Classification",
       y = "Number of Incidents") +

  coord_flip()

# plot
print(g)


```

# Modeling Factors that can Contribute to the Fatality of a Shooting

```{r}


NYPD$STATISTICAL_MURDER_FLAG <- as.integer(NYPD$STATISTICAL_MURDER_FLAG == "true")

# Fit the model with all specified variables
glm.fit <- glm(STATISTICAL_MURDER_FLAG ~ PERP_RACE + PERP_SEX + PERP_AGE_GROUP + 
               VIC_AGE_GROUP + VIC_SEX + VIC_RACE + BORO + HOUR_OF_DAY + 
               DAY_OF_WEEK + 
               LOC_CLASSFCTN_DESC + LOCATION_DESC + LOC_OF_OCCUR_DESC + 
               PRECINCT, data = NYPD, family = binomial)


# Summary of the model
summary(glm.fit)

# Obtain tidy version of the glm model
tidy_glm <- tidy(glm.fit)

# Remove NA coefficients and sort the coefficients in descending order
sorted_tidy_glm <- tidy_glm %>%
  filter(!is.na(estimate)) %>%
  arrange(desc(estimate))

# Print the sorted table of coefficients
print(sorted_tidy_glm)


```

```{r}

tidy_glm <- tidy(glm.fit)

# Remove rows with NA coefficients, non-significant p-values, and the intercept
tidy_glm <- tidy_glm %>%
  filter(!is.na(estimate) & p.value < 0.001 & term != "(Intercept)")

# Create a plot 
ggplot(tidy_glm, aes(x = reorder(term, estimate), y = estimate)) +
  geom_col(fill = "#3BB9FF", color = "black", show.legend = FALSE) +  
  coord_flip() +  
  labs(x = "Coefficient", y = "Predictor", title = "Significant Coefficients of Predictors") +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold", hjust = 3),  # Align title to the left properly
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.margin = margin(10, 10, 10, 10)  # Adjust plot margins if needed
  )

```

\`\`\`

# Conclusion

-   **NYC Incident Trends**: Declining incidents with a sharp drop present, possibly due to global factors.

-   **Day of the Week**: Increased incidents on weekends

-   **Hourly Distribution**: most incidents during evening hours; fewest in the early morning.

-   **Borough Analysis**: Higher incident rates in Brooklyn and the Bronx; Manhattan has fewer murder-classified incidents.

-   **Location**: Incidents are most common in public housing units

**Factors Which Contribute to Fatality - Logistic Regression Insights:**

-   Race: Specific racial groups such as White, Asian/Pacific Islander, Black, and Black Hispanic have higher odds of incidents being classified as murders, underscoring the influence of racial dynamics on shooting outcomes.

-   Location Type: Video stores and drug stores, as indoor locations, show strong associations with murder classifications, likely due to the close-range nature of incidents in these settings.

-   Unknown Age Group: The 'UNKNOWN' age category for perpetrators is significantly associated with fatal outcomes, indicating challenges or risks associated with unidentified offenders.

-   Victim Age: Older victims, particularly those over 65, are more frequently involved in fatal shootings, pointing to their increased vulnerability.

-   Perpetrator Gender: The presence of an "UNKNOWN" category for gender highlights potential ambiguities in the data, affecting the analysis of gender impacts on murder classifications.

Summary: The analysis reveals significant racial and locational factors in murder classifications within NYPD shooting incidents, with additional complexity introduced by unknown demographic data. These findings highlight the critical areas for law enforcement focus and further research to address and mitigate the risks associated with these key predictors.

-   **Potential Biases**:

    -   Data Collection Bias: Risks due to non-representative data or reporting variances.

    -   Measurement Bias: Inconsistencies in incident classification processes.

    -   Omitted Variable Bias: Missing variables could distort the analysis.

    -   Modeling Bias: The model may not accurately capture the complexity of incident classifications.
